---
title: "SAB01 Data Prep"
author: "Shalise Burch"
output: html_document
date: "2024-11-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#ONLY NEED TO RUN THIS RMD ONCE, FOLLOW UP RMDS USE SAVED TIDY DATA
RMD RAN: 11.11.24 @ 9:18PM

#DATA = SAB01 flow data
SET 3 D21 and T21 - with 2 treatment arms RA and CTRL


## Data Prep :
STEP1: Import data and Prep Data
STEP2: Visually Inspect Data 
STEP3: Prep Tools for Downstream Analysis


#STEP1: Import and Check Data 
```{r load and check data}

library(readxl)

#import data 
data <- read_excel("00_raw_data/r.SAB01flow.xlsx", 
    sheet = "r.SAB01")
data


```


```{r convert decimals to percent values}

# check data to make sure you need this step, excel often saves data as percentages in decimal form 
data$dbl_neg <- data$dbl_neg * 100
data$isl_pos <- data$isl_pos * 100
data$islctnt_pos <- data$islctnt_pos * 100
data$tot_isl <- data$tot_isl * 100
data$ctnt_pos <- data$ctnt_pos * 100
data$tot_ctnt <- data$tot_ctnt * 100

data

```

```{r save transformed data}
write.csv(data,"01_tidy_data/data.csv", row.names = FALSE)
```

#STEP2: Visually Inspect Data -- total islet1 and total ctnt only 
```{r data visualization} 
library(dplyr)
library(tidyverse)
library(ggpubr)
library(ggthemes)

#option1
qplot(x = day, y = tot_isl, geom = "point", color = treatment, shape=treatment, size=1, data = data) +
  facet_grid(.~cell, labeller = label_both)


qplot(x = day, y = tot_ctnt, geom = "point", color = treatment, shape=treatment, size=1, data = data) +
  facet_grid(.~cell, labeller = label_both)


#option2 
ggboxplot(data, x = "day", y = "tot_isl", color = "cell", ylab = "Expression", xlab = "Day")+ 
  facet_grid(.~treatment, labeller = label_both)


ggboxplot(data, x = "day", y = "tot_ctnt", color = "cell", ylab = "Expression", xlab = "Day")+
    facet_grid(.~treatment, labeller = label_both)

#can save plot if desired -- this code will save last generated plot -- 
#ggsave(path = "03_plots", filename = str_c("qplot",  format(Sys.Date(), "%Y%m%d"), format(Sys.time(), "_%H%M"), ".pdf"), height = 4, width = 8, units = "in")
```

#STEP3 Create tools to use in down stream analysis
UPDATE AS NEEDED

```{r measurments}

#Vector of flow measurements to use in loops, also used to order graph facet so put in specific order 

measurements <- c("dbl_neg", "isl_pos", "islctnt_pos", "ctnt_pos", "tot_isl", "tot_ctnt")
saveRDS(measurements, file = "01_tidy_data/measurements.rds") #save as R object not CVS

```


```{r long from }

# Reshape flow measurement columns to long form for looping and applying raw data points to bar graphs. 
long_data <- data %>%
  pivot_longer(cols = c(all_of(measurements)),  #only reshape the measurements you specified above 
               names_to = "measurement",         #column for measurement names
               values_to = "value")              #column for measurement values


write.csv(long_data,"01_tidy_data/long_data.csv", row.names = FALSE )
```


```{r color customization}

# Create data frame specifying colors for each treatment and cell type to use in graphs
color_map <- data.frame(
  Group = c("CTRL", "RA", "VIS", "SAG", "RA&SAG", "RA&VIS"),
  Color = c("orange", "red", "darkorchid1", "cornflowerblue", "darkturquoise", "chartreuse2")
)

write.csv(color_map,"01_tidy_data/color_map.csv", row.names = FALSE)
```


```{r define treatment order for graphing}

treatment_order <- c("CTRL", "SAG", "VIS", "RA", "RA&SAG", "RA&VIS")
saveRDS(treatment_order, file = "01_tidy_data/treatment_order.rds") #save as R object not CVS

```


##DONE. Don't need to run this RMD again--- will used tidy data for downstream analysis 

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
