---
title: "First Pass Data Visulization- SAB01"
output: html_document
date: "2024-11-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Inspection and Summary Stats:
STEP1: Load Tidy Data 
STEP2: Generate Summary Statistics 
STEP3: Generate First Pass Bar Graphs of Data


#STEP1: Load All Data Needed
```{r load data, message=FALSE, warning=FALSE}

library(readr)
data <- read_csv("01_tidy_data/data.csv")
long_data <- read_csv("01_tidy_data/long_data.csv")
color_map <- read_csv("01_tidy_data/color_map.csv")
measurements <- readRDS("01_tidy_data/measurements.rds")
treatment_order <- readRDS("01_tidy_data/treatment_order.rds")

```

#STEP2: Summary Statistics
Loop through each measurement and produce summary stats for each

```{r sumstats}

library(dplyr)
library(tidyverse)


results_list <- list() #empty list to store results


for (measurement in measurements) {
  summary_data <- data %>%
    group_by(cell, day, treatment) %>%
    summarise(mean = mean(get(measurement), na.rm = TRUE), 
              sd = sd(get(measurement), na.rm = TRUE)) %>%
    arrange(desc(mean))
  
  summary_data <- summary_data %>%
    mutate(measurement = measurement) #creates column for measurement 
  
  # Store the result in  list with name corresponding to measure
  results_list[[measurement]] <- summary_data
}

sumstats_all <- bind_rows(results_list) #combine into one dataframe
print(sumstats_all)

# save
write.csv(sumstats_all,"02_tables/sumstats_all.csv", row.names = FALSE)


```




#STEP3: DATA VISUALIZATION
* Run theme_publication in theme_Publication.R *
Data needed: sumstats and raw data which is already loaded 

```{r all measurments }

library(ggplot2)
library(ggpubr)
library(ggthemes)

# Convert measurement and treatment to a factor with the specified order -- to be used in the facet
sumstats_all$measurement <- factor(sumstats_all$measurement, levels = measurements)
sumstats_all$treatment <- factor(sumstats_all$treatment, levels = treatment_order)


bargraph.all<- ggplot(sumstats_all, aes(x = interaction(day), y = mean, fill = cell)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(measurement ~ treatment, scales = "free_y") + 
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, position = position_dodge(0.9)) +
  labs(title = "Bar Graph of Mean Values for Each Measurement",
       x = "Day",
       y = NULL,
       fill = "Cell Type") +
  theme_Publication() +
  theme(legend.position = "top",
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 11),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14))

bargraph.all

```


```{r save}
ggsave(bargraph.all, path = "03_plots",
      filename = str_c("all.bargraph.pdf"),
       height = 11, width = 10, units = "in")
```

#Generate graphs
Plots for each measurement:

(2) FACETED graphs
(a) by treatment to compare T21vD21 
(b) by cell to compare treatment 
```{r create subfolder for individual treatment graphs}
# Define the path and folder name
folder_path <- "03_plots/faceted"

# Create the folder
if (!dir.exists(folder_path)) {
  dir.create(folder_path, recursive = TRUE)
  message("Folder created at: ", folder_path)
} else {
  message("Folder already exists at: ", folder_path)
}
```

```{r generating facet graphs}

for (measure in measurements) {
  measurement_data <- sumstats_all %>% 
    filter(measurement == measure)
  
  raw_data_subset <- long_data %>%
    filter(measurement == measure)
  
  # bar graph facet by treatment comparing cell type 
  p <- ggplot(measurement_data, aes(x = factor(day), y = mean, fill = treatment, color= treatment)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.8), 
           width = 0.7, size = 1, aes(alpha = cell, group = interaction(cell, treatment))) +
    scale_y_continuous(limits = c(0, 100), labels = scales::percent_format(,1)) +
    scale_fill_manual(name = "S3D21", values = setNames(color_map$Color, color_map$Group)) +
    scale_color_manual(name = "S3T21", values = setNames(color_map$Color, color_map$Group)) +
    scale_alpha_manual(values = c("S3D21" = 0.9, "S3T21" = 0)) +
    geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd, group = interaction(cell, treatment)), 
                width = 0.4, size = 0.5, position = position_dodge(width = 0.8), color = "black") +
    geom_point(data = raw_data_subset, aes(x = factor(day), y = value, group = interaction(cell, treatment)), 
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), 
             size = 1.5, alpha = 0.8, color = "black", show.legend = FALSE) +
    
    facet_wrap(~factor(treatment, levels = treatment_order))+ # Facet by treatment, specify order 
    labs(title = paste(measure, "- Comparing Cell Type within Treatment"),
         x = "Day",
         y = measure) +
   theme_Publication() +
   guides(
    fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
    color = guide_legend(order = 2, override.aes = list(fill = NA)), 
    alpha = "none"  # Hide alpha legend
  ) 
  
  
   # bar graph facet by cell comparing treatment
  p2 <- ggplot(measurement_data, aes(x = factor(day), y = mean, fill = treatment, color= treatment)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.8), 
           width = 0.7, size = 1, aes(alpha = cell, group = interaction(cell, treatment))) +
    scale_y_continuous(limits = c(0, 100), labels = scales::percent_format(,1)) +
    scale_fill_manual(name = "S3D21", values = setNames(color_map$Color, color_map$Group)) +
    scale_color_manual(name = "S3T21", values = setNames(color_map$Color, color_map$Group)) +
    scale_alpha_manual(values = c("S3D21" = 0.9, "S3T21" = 0)) +
    geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd, group = interaction(cell, treatment)), 
                width = 0.4, size = 0.5, position = position_dodge(width = 0.8), color = "black") +
    geom_point(data = raw_data_subset, aes(x = factor(day), y = value, group = interaction(cell, treatment)), 
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), 
             size = 1.5, alpha = 0.8, color = "black", show.legend = FALSE) +
     
    facet_wrap(~cell)+ 
    labs(title = paste(measure, "-  Comparing Treatment within Cell Type"),
         x = "Day",
         y = measure,
         fill = "Treatment") +
    theme_Publication() +
   guides(
    fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
    color = guide_legend(order = 2, override.aes = list(fill = NA)), 
    alpha = "none"  # Hide alpha legend
  ) 
   
  
  # Save plots
  ggsave(filename = paste0(measure, "_facet_treatment.pdf"), path = "03_plots/faceted",
         plot = p, width = 10, height = 8, units = "in")
  
   ggsave(filename = paste0(measure, "_facet_cell.pdf"), path = "03_plots/faceted",
         plot = p2, width = 10, height = 8, units = "in")
}
  

#time stamps
#  ggsave(filename = paste0(measure, "_facet_treatment", format(Sys.Date(), "%Y%m%d"), format(Sys.time(), "-%H%M"), ".pdf"), path = "03_plots/faceted",
 #        plot = p, width = 10, height = 8, units = "in")

```


(3)By treatment graphs 
```{r create subfolder for individual treatment graphs}
# Define the path and folder name
folder_path <- "03_plots/bytreatment"

# Create the folder
if (!dir.exists(folder_path)) {
  dir.create(folder_path, recursive = TRUE)
  message("Folder created at: ", folder_path)
} else {
  message("Folder already exists at: ", folder_path)
}
```

```{r by treatment graphs}


for (measure in measurements) {
  summary_data <- sumstats_all %>% 
    filter(measurement == measure)
  
  measure_data <- long_data %>%
    filter(measurement == measure)
  
  # Loop through each treatment to plot individually
  for (treatment_val in unique(measure_data$treatment)) {
    
    # Filter data for the current treatment
    treatment_data <- measure_data %>% filter(treatment == treatment_val)
    treatment_summary <- summary_data %>% filter(treatment == treatment_val)
    
    # Plot comparing cell types within each treatment, with days on x-axis and y as the measurement
    p <- ggplot(treatment_summary, aes(x = factor(day), y = mean, fill = treatment, color = treatment)) +
     # Bars with consistent position dodge and group interaction
    geom_bar(stat = "identity", position = position_dodge(width = 0.8), 
           width = 0.8, size = 1, aes(alpha = cell, group = interaction(cell, treatment))) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20), labels = scales::percent_format(,1)) +
  
  # Custom colors for fill (cell type) and outline (treatment)
    scale_fill_manual(name = "S3D21", values = setNames(color_map$Color, color_map$Group)) +
    scale_color_manual(name = "S3T21", values = setNames(color_map$Color, color_map$Group)) +
    scale_alpha_manual(values = c("S3D21" = 0.9, "S3T21" = 0)) +

  # Error bars with the same position dodge and group interaction
    geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd, group = interaction(cell, treatment)), 
                width = 0.4, size = 0.5, position = position_dodge(width = 0.8), color = "black") +

  # Data points for individual measurements with the same group interaction
    geom_point(data = treatment_data, aes(x = factor(day), y = value, group = interaction(cell, treatment)), 
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), 
             size = 1.5, alpha = 0.8, color = "black", show.legend = FALSE) +

    labs(
      title = paste(measure, "-", treatment_val, "treatment"),
      x = "Day", 
      y = measure) +

   theme_Publication() +
   guides(
    fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
    color = guide_legend(order = 2, override.aes = list(fill = NA)), 
    alpha = "none"  # Hide alpha legend 
  ) 
    
    # Define the filename for saving the plot
    file_name <- paste0(treatment_val, "_", measure, ".pdf")
  
    ggsave(file_name, path = "03_plots/bytreatment", plot = p, width = 7, height = 6, units = "in")
  
  }
}
```


(4)By cell graphs
```{r create subfolder for individual cell graphs}
# Define the path and folder name
folder_path <- "03_plots/bycell"

# Create the folder
if (!dir.exists(folder_path)) {
  dir.create(folder_path, recursive = TRUE)
  message("Folder created at: ", folder_path)
} else {
  message("Folder already exists at: ", folder_path)
}
```

```{r by cell graphs}

for (measure in measurements) {
  summary_data <- sumstats_all %>% 
    filter(measurement == measure)
  
  measure_data <- long_data %>%
    filter(measurement == measure)
  
  # Loop through each cell type to plot individually
  for (cell_type in unique(measure_data$cell)) {
    
    # Filter data for the current cell type
    cell_data <- measure_data %>% filter(cell == cell_type)
    cell_summary <- summary_data %>% filter(cell == cell_type)
    
    # Check if specific treatments are present in the filtered data
    has_S3D21 <- "S3D21" %in% unique(cell_summary$cell)
    has_S3T21 <- "S3T21" %in% unique(cell_summary$cell)
    
    # Start building the ggplot
    p <- ggplot(cell_summary, aes(x = factor(day), y = mean, fill = treatment, color = treatment)) +
      geom_bar(stat = "identity", position = position_dodge(width = 0.9), 
               width = 0.8, size = 1, aes(alpha = cell, group = interaction(cell, treatment))) +
      scale_y_continuous(limits = c(0, 100), labels = scales::percent_format(,1)) +
      scale_fill_manual(name = "Treatment", values = setNames(color_map$Color, color_map$Group), drop = TRUE) +
      scale_color_manual(name = "Treatment", values = setNames(color_map$Color, color_map$Group), drop = TRUE) +
      scale_alpha_manual(values = c("S3D21" = 0.9, "S3T21" = 0)) +
      geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd, group = interaction(cell, treatment)), 
                    width = 0.3, size = 0.5, position = position_dodge(width = 0.9), color = "black") +
      geom_point(data = cell_data, aes(x = factor(day), y = value, group = interaction(cell, treatment)), 
                 position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.9), 
                 size = 1.5, alpha = 0.8, color = "black", show.legend = FALSE) +
      labs(
        title = paste(measure, "-", cell_type),
        x = "Day", 
        y = measure
      ) +
      theme_Publication() 
    
    # Conditional legends based on presence of treatments
    guides_list <- list(alpha = "none")  # Always hide alpha legend
    
    if (has_S3D21) {
      guides_list$fill <- guide_legend(override.aes = list(alpha = 1))
      guides_list$color <- "none"  # Hide color legend if S3D21 is present
}

    if (has_S3T21) {
      guides_list$color <- guide_legend(override.aes = list(fill = NA))
      guides_list$fill <- "none"  
    }
    
    p <- p + do.call(guides, guides_list)
    
    # Save the plot
    file_name <- paste0(cell_type, "_", measure, ".pdf")
    ggsave(file_name, path = "03_plots/bycell", plot = p, width = 5, height = 6, units = "in")
  }
}
```

(5) By day - all measurements  
```{r create subfolder for individual treatment graphs}
# Define the path and folder name
folder_path <- "03_plots/byday"

# Create the folder
if (!dir.exists(folder_path)) {
  dir.create(folder_path, recursive = TRUE)
  message("Folder created at: ", folder_path)
} else {
  message("Folder already exists at: ", folder_path)
}
```

```{r by day graphs}


for (current_day in unique(long_data$day)) {
  
  # Filter data for the current day
  summary_data <- sumstats_all %>%
    filter(day == current_day)
  
  day_data <- long_data %>%
    filter(day == current_day)
  
  # Loop through each treatment to plot individually
  for (treatment_val in unique(day_data$treatment)) {
    
    # Filter data for the current treatment
    treatment_data <- day_data %>%
      filter(treatment == treatment_val)
    treatment_summary <- summary_data %>%
      filter(treatment == treatment_val)
    
    # Plot comparing measurements within each treatment, with measurements on x-axis and y as the mean value
    p <- ggplot(treatment_summary, aes(x = factor(measurement), y = mean, fill = treatment_val, color = treatment_val)) +
      # Bars with consistent position dodge and group interaction
      geom_bar(stat = "identity", position = position_dodge(width = 0.8), 
               width = 0.8, size = 1, aes(alpha = cell, group = interaction(cell, measurement))) +
      scale_y_continuous(labels = scales::percent_format(,1)) +
      
      # Custom colors for fill (cell type) and outline (cell type)
       scale_fill_manual(name = "S3D21", values = setNames(color_map$Color, color_map$Group)) +
    scale_color_manual(name = "S3T21", values = setNames(color_map$Color, color_map$Group)) +
    scale_alpha_manual(values = c("S3D21" = 0.9, "S3T21" = 0)) +
      
      # Error bars with the same position dodge and group interaction
      geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd, group = interaction(cell, measurement)), 
                    width = 0.4, size = 0.5, position = position_dodge(width = 0.8), color = "black") +
      
      # Data points for individual measurements with the same group interaction
      geom_point(data = treatment_data, aes(x = factor(measurement), y = value, group = interaction(cell, measurement)), 
                 position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), 
                 size = 1.5, alpha = 0.8, color = "black", show.legend = FALSE) +
      
      labs(
        title = paste("Day", current_day, "-", treatment_val, "treatment"),
        x = NULL, 
        y = NULL
      ) +
      
      theme_Publication() +
      #theme(legend.position = "right")+
      guides(
      fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
      color = guide_legend(order = 2, override.aes = list(fill = NA)), 
      alpha = "none"  # Hide alpha legend 
    ) 
    
    # Define the filename for saving the plot
    file_name <- paste0("Day_", current_day, "_", treatment_val, ".pdf")
    
    ggsave(file_name, path = "03_plots/byday", plot = p, width = 6, height = 6, units = "in")
  }
}
```



(6) COMBINED GRAPH - D21 + T21 multiple treatments 
```{r filter}
#filtering data for days across treatment, not always necessary but for this data Ra was only collected on even days and ctrl was collected all days 

sumstats_all <- sumstats_all %>% filter(day %in% c("4", "6", "8", "10"))
long_data <- long_data %>% filter(day %in% c("4", "6", "8", "10"))


```

#there are two treatments present CTRL and RA. 
```{r generating graphs}

for (measure in measurements) {
  measurement_data <- sumstats_all %>% 
    filter(measurement == measure)
  
  raw_data_subset <- long_data %>%
    filter(measurement == measure)
  
  # bar graph grouped by treatment comparing cell type 
  p <- ggplot(measurement_data, aes(x = factor(day), y = mean, fill = treatment, color = treatment)) +
  
    # Bars with consistent position dodge and group interaction
    geom_bar(stat = "identity", position = position_dodge(width = 0.8), 
           width = 0.7, size = 1, aes(alpha = cell, group = interaction(cell, treatment))) +
    scale_y_continuous(limits = c(0, 100), labels = scales::percent_format(,1)) +
  
  # Custom colors for fill (cell type) and outline (treatment)
    scale_fill_manual(name = "S3D21", values = setNames(color_map$Color, color_map$Group)) +
    scale_color_manual(name = "S3T21", values = setNames(color_map$Color, color_map$Group)) +
    scale_alpha_manual(values = c("S3D21" = 0.9, "S3T21" = 0)) +

  # Error bars with the same position dodge and group interaction
    geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd, group = interaction(cell, treatment)), 
                width = 0.4, size = 0.5, position = position_dodge(width = 0.8), color = "black") +

  # Data points for individual measurements with the same group interaction
    geom_point(data = raw_data_subset, aes(x = factor(day), y = value, group = interaction(cell, treatment)), 
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), 
             size = 1.5, alpha = 0.8, color = "black", show.legend = FALSE) +

    labs(x = "Day", 
       y = measure) +

   theme_Publication() +
   guides(
    fill = guide_legend(order = 1, override.aes = list(alpha = 1)),
    color = guide_legend(order = 2, override.aes = list(fill = NA)), 
    alpha = "none"  # Hide alpha legend
  ) +

  theme(legend.position = "right")

  
  # Save plots
  ggsave(filename = paste0(measure, "_combined.pdf"), path = "03_plots",
         plot = p, width = 10, height = 8, units = "in")
  
}
  
```

```{r xyz}

```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
