---
title: "Assumptions and Statistical Comparisons -- TRT vs CTRL"
output: html_document
date: "2024-11-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Analysis overview
For each day and each measurement:
    (1) Compare TRTvCTRL for each cell type 
    (2) Compare T21vD21 within each treatment 

## Data Inspection and Summary Stats:
STEP1: Load Libraries and Data 
STEP2: Check Assumptions
STEP3: Run Statistical Analysis  
STEP4: Generate Matching Bar Plots 


STEP1: Load Libraries and Data 
```{r loading libraries, include=FALSE}

library(readr)
library(dplyr)
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(ggplot2)

```

```{r load data, include=FALSE}

data <- read_csv("01_tidy_data/data.csv")
long_data <- read_csv("01_tidy_data/long_data.csv")
color_map <- read_csv("01_tidy_data/color_map.csv")
measurements <- readRDS("01_tidy_data/measurements.rds")
treatment_order <- readRDS("01_tidy_data/treatment_order.rds")
sumstats_all <- read_csv("02_tables/sumstats_all.csv") 

```

# STEP2: Check  Assumptions for Data in Comparison 

TRTvCTRL  and D21vT21 
```{r variance and normality }
library(car) #for levenes test

# Data frame to store results
assumption_results <- data.frame(
  Day = integer(),
  Measurement = character(),
  Treatment1 = character(),
  Treatment2 = character(),
  Cell1 = character(),
  Cell2 = character(),
  Comparison = character(),
  Control_Shapiro_PValue = numeric(),
  Treatment_Shapiro_PValue = numeric(),
  Normality = character(),
  Levene_PValue = numeric(),
  Variance_equal = logical(),
  stringsAsFactors = FALSE
)

# Specify  control 
control_label <- "CTRL" 

# Loop through days and measurement to run tests
for (day_val in unique(data$day)) {
  for (measure in measurements) {
    
    # Filter data for the current day and measurement
    day_data <- data %>%
      filter(day == day_val) %>%
      select(cell, treatment, !!sym(measure)) %>%
      na.omit() 
    
    #Using day data loop through cell type to compare each treatment to control
    for (cell_type in unique(day_data$cell)) {
      cell_data <- day_data %>% filter(cell == cell_type)
      
      # Filter control
      control_data <- cell_data %>% filter(treatment == control_label)
      
      # Within each cell type loop through treatments for comparing against control
      for (treatment_val in unique(cell_data$treatment)) {
        if (treatment_val != control_label) {
          
          # Filter for treatment group 
          treatment_data <- cell_data %>% filter(treatment == treatment_val)
          
          # Combine control and treatment data
          comparison_data <- bind_rows(control_data, treatment_data)
          
          # Run tests if at least two groups (control and  treatment)
          if (n_distinct(comparison_data$treatment) >= 2) {
            
            # Shapiro-Wilk test for normality within each group
            control_shapiro <- shapiro.test(control_data[[measure]])$p.value
            treatment_shapiro <- shapiro.test(treatment_data[[measure]])$p.value
            normality <- ifelse(control_shapiro > 0.05 & treatment_shapiro > 0.05, "Normal", "Not Normal")
            
            # Levene’s test for homogeneity of variance 
            levene_test <- leveneTest(as.formula(paste(measure, "~ treatment")), data = comparison_data)
            variance_equal <- levene_test$`Pr(>F)`[1] > 0.05  # TRUE if variances are equal
            
            # Store results for normality and variance tests in a single entry
            assumption_results <- assumption_results %>%
              bind_rows(data.frame(
                Day = day_val,
                Measurement = measure,
                Treatment1 = control_label,
                Treatment2 = treatment_val,
                Cell1 = cell_type,
                Cell2 = "NA",
                Comparison = paste0(cell_type, "--", treatment_val, "vs", control_label),
                Control_Shapiro_PValue = control_shapiro,
                Treatment_Shapiro_PValue = treatment_shapiro,
                Normality = normality,
                Levene_PValue = levene_test$`Pr(>F)`[1],
                Variance_equal = variance_equal
              ))
          }
        }
      }
    }
    
    # Loop through treatments to compare cell types within that treatment
    for (treatment_val in unique(day_data$treatment)) {
      treatment_data <- day_data %>% filter(treatment == treatment_val)
      
      # Run tests if there are at least two unique cell types
      if (n_distinct(treatment_data$cell) >= 2) {
        
        # designate one cell as "control" and the other as "treatment" for storing data 
        control_cell <- "S3D21"
        treatment_cell <- "S3T21"
        
        # Filter data for the control and treatment cell types
        control_data <- treatment_data %>% filter(cell == control_cell)
        treatment_data <- treatment_data %>% filter(cell == treatment_cell)
        
        # Combine control and treatment data
        comparison_data <- bind_rows(control_data, treatment_data)
          
        control_shapiro <- shapiro.test(control_data[[measure]])$p.value
        treatment_shapiro <- shapiro.test(treatment_data[[measure]])$p.value
        normality <- ifelse(control_shapiro > 0.05 & treatment_shapiro > 0.05, "Normal", "Not Normal")
   
        levene_test <- leveneTest(as.formula(paste(measure, "~ cell")), data = comparison_data)
        variance_equal <- levene_test$`Pr(>F)`[1] > 0.05  # TRUE if variances are equal

        assumption_results <- assumption_results %>%
          bind_rows(data.frame(
            Day = day_val,
            Measurement = measure,
            Treatment1 = treatment_val,
            Treatment2 = "NA",
            Cell1 = control_cell,
            Cell2 = treatment_cell,
            Comparison = paste0(treatment_val, "--", treatment_cell, "vs", control_cell),
            Control_Shapiro_PValue = control_shapiro,  
            Normality = normality,
            Treatment_Shapiro_PValue = treatment_shapiro,
            Levene_PValue = levene_test$`Pr(>F)`[1],
            Variance_equal = variance_equal
          ))
      }
    }
  }
}

write.csv(assumption_results,"02_tables/assumption_results.csv", row.names = FALSE)
print(assumption_results)

```

# STEP3: Run Statistical Analysis
Common tests in one loop

```{r comparision }
library(ggpubr)

results_cell <- data.frame(
  Measurement = character(),
  Day = numeric(),
  Treatment = character(),
  Paired_TTest_PValue = numeric(),
  Welch_TTest_PValue = numeric(),
  Kruskal_Wallis_PValue = numeric(),
  Wilcoxon_PValue = numeric(),
  stringsAsFactors = FALSE
)

results_treatment <- data.frame(
  Measurement = character(),
  Day = numeric(),
  Cell = character(),
  Compared_Treatment = character(),  
  Paired_TTest_PValue = numeric(),
  Welch_TTest_PValue = numeric(),
  Kruskal_Wallis_PValue = numeric(),
  Wilcoxon_PValue = numeric(),
  stringsAsFactors = FALSE
)

# Outer loop: Compare each treatment to the control group
control_group <- "CTRL"  # Define control group 
treatment_groups <- unique(data$treatment[data$treatment != control_group])

for (treatment_val in treatment_groups) {
  # Subset data for the current treatment and control
  subset_data <- data %>%
    filter(treatment %in% c(control_group, treatment_val))

  # Inner Loop 1: Compare cell types within a treatment and day for each measurement
  for (measure in measurements) {
    for (day_val in unique(subset_data$day)) {
      for (current_treatment in unique(subset_data$treatment)) {
       
         # Filter data for current day and  treatment
        data_day_treatment <- subset_data %>%
          filter(day == day_val, treatment == current_treatment)

        # Perform tests if there are at least two cell types
        if (length(unique(data_day_treatment$cell)) > 1) {
          kruskal_test <- kruskal.test(as.formula(paste(measure, "~ cell")), data = data_day_treatment)
          
          
          # Run Wilcoxon, Welch, and Paired t-tests if exactly two cell types
          if (length(unique(data_day_treatment$cell)) == 2) {
            
            wilcox_test <- wilcox.test(as.formula(paste(measure, "~ cell")), data = data_day_treatment)
            wilcox_p <- wilcox_test$p.value
            
            welch_test <- t.test(as.formula(paste(measure, "~ cell")), data = data_day_treatment, var.equal = FALSE)
            welch_p <- welch_test$p.value
            
            # Paired t-test 
            # Step 1: Reshape the data to align CTRL and RA replicates
            data_paired <- data_day_treatment %>%
              select(exp_no, cell, measure) %>%  
              spread(key = cell, value = measure)
            
            # Step 3: Perform a paired t-test
            if (all(c("S3D21", "S3T21") %in% colnames(data_paired))) {
              paired_test <- t.test(data_paired$S3D21, data_paired$S3T21, paired = TRUE)
              paired_p <- paired_test$p.value
            } 
          }
        
          
          # Store results for cell type comparisons
          results_cell <- rbind(results_cell, data.frame(
            Measurement = measure,
            Day = day_val,
            Treatment = current_treatment,
            Paired_TTest_PValue = paired_p,
            Welch_TTest_PValue = welch_p,
            Kruskal_Wallis_PValue = kruskal_test$p.value,
            Wilcoxon_PValue = wilcox_p
          ))
        }
      }
    }
  }
  
  # Inner Loop 2: Compare treatments within each cell type and day for each measurement
  for (measure in measurements) {
    for (day_val in unique(subset_data$day)) {
      for (cell_val in unique(subset_data$cell)) {
        
        # Filter data for current day, cell type, and measurement
        data_day_cell <- subset_data %>%
          filter(day == day_val, cell == cell_val)
        
        # Perform tests only if there are at least two treatments
        if (length(unique(data_day_cell$treatment)) > 1) {
          kruskal_test <- kruskal.test(as.formula(paste(measure, "~ treatment")), data = data_day_cell)
          
          # Run Wilcoxon, Welch, and Paired t-tests if exactly two treatments
          if (length(unique(data_day_cell$treatment)) == 2) {
            
            wilcox_test <- wilcox.test(as.formula(paste(measure, "~ treatment")), data = data_day_cell)
            wilcox_p <- wilcox_test$p.value
            
            welch_test <- t.test(as.formula(paste(measure, "~ treatment")), data = data_day_cell, var.equal = FALSE)
            welch_p <- welch_test$p.value
            
            # Paired t-test 
            # Step 1: Reshape the data 
            data_paired <- data_day_cell %>%
              select(exp_no, treatment, measure) %>%  
              spread(key = treatment, value = measure)
            
            # Step 3: Perform a paired t-test
            if (all(c(control_group, treatment_val) %in% colnames(data_paired))) {
              paired_test <- t.test(data_paired[[control_group]], data_paired[[treatment_val]], paired = TRUE)
              paired_p <- paired_test$p.value
            } 
          }
          
          # Store results
          results_treatment <- rbind(results_treatment, data.frame(
            Measurement = measure,
            Day = day_val,
            Cell = cell_val,
            Compared_Treatment = paste(treatment_val, "vs", control_group),
            Paired_TTest_PValue = paired_p,
            Welch_TTest_PValue = welch_p,
            Kruskal_Wallis_PValue = kruskal_test$p.value,
            Wilcoxon_PValue = wilcox_p
            
          ))
        }
      }
    }
  }
}


print(results_cell)
print(results_treatment)



```

```{r save stats }
write.csv(results_cell,"02_tables/cell_stats.csv")
write.csv(results_treatment,"02_tables/treatment_stats.csv")
```

```{r xyz }

```



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
