---
title: "Cell Growth Analysis"
output: html_document
date: "2024-11-8"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Analy1sis pipeline:
STEP1: Import and Check Data 
STEP2: Summary Stats and Graphs
STEP3: Check Assumptions - Variance and Normality 
STEP4: Run Appropriate Stat Tests




```{r load libraries}

library(readxl)
library(dplyr)
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(ggplot2)

```

#STEP1: Import and Check Data 
```{r load and check data}

count_data<- read_excel("00_raw_data/r.SAB01.counts.xlsx", 
    sheet = "r.SAB01")
count_data

treatment_order <- readRDS("01_tidy_data/treatment_order.rds")
color_map <- read_csv("01_tidy_data/color_map.csv")

count_data$day <- factor(count_data$day, levels = c("-2", "3", "4", "5", "6", "7", "8", "9", "10")) 


```



```{r data visualization} 

boxplot<-ggboxplot(count_data, x = "day", y = "cell_cm2", 
          color = "cell", 
          ylab = "Cell per Cm2", xlab = "Day")+
    facet_grid(.~treatment, labeller = label_both)
boxplot

```



#STEP2: Summary Statistics
```{r summary statistic}

counts_sumstats <- group_by(count_data, cell, day, treatment) %>%
  summarise(mean=mean(cell_cm2), sd=sd(cell_cm2)) %>%
  arrange(desc(mean))

write.csv(counts_sumstats,"02_tables/counts_sumstats.csv")
print(counts_sumstats)


```



#GRAPHS 

```{r grid overview of data }

line.grid<- ggplot(counts_sumstats, aes(x=day, y=mean, color=treatment, group=treatment)) +
  scale_color_manual(values = setNames(color_map$Color, color_map$Group))+
  scale_y_continuous(limits = c(0, 1200000), breaks = seq(0, 12000000, 200000),label=scales::comma)+
  geom_line(size=1)+
  geom_point()+
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.5, size=1,
                 position=position_dodge(0.05))+
  theme_bw() +
  labs(
  title = "Cell Growth Over Time",
  x = "Day",
  y = "Cells per Cm2",
  fill = "")+
  facet_grid(cell ~ treatment)
line.grid



linebf.grid<- ggplot(count_data, aes(x=day, y=cell_cm2, fill=treatment, group=treatment))+
  scale_fill_manual(values = setNames(color_map$Color, color_map$Group))+
  scale_y_continuous(limits = c(0, 1200000), breaks = seq(0, 12000000, 200000),label=scales::comma)+
  geom_point(shape=21) +
  geom_smooth(method = "loess",
              color="black",
              size=0.5)+
  theme_bw() +
  labs(
  title = "Cell Growth Over Time",
  x = "Day",
  y = "Cells per Cm2",
  fill = "")+
  facet_grid(cell ~ treatment)
linebf.grid

```

```{r save }
ggsave(linebf.grid, path = "03_plots/cell_growth",
      filename = str_c("linebf_grid_",  format(Sys.Date(), "%Y%m%d"), format(Sys.time(), "-%H%M"), ".pdf"),
       height = 4, width = 10, units = "in")

ggsave(line.grid, path = "03_plots/cell_growth",
      filename = str_c("line_grid_",  format(Sys.Date(), "%Y%m%d"), format(Sys.time(), "-%H%M"), ".pdf"),
       height = 4, width = 10, units = "in")
```

```{r bar graph}

#Separated by treatment
bar1<-ggplot(counts_sumstats, aes(x=day, y=mean, fill=cell)) + 
  scale_fill_manual(values = setNames(color_map$Color, color_map$Group)) +
  scale_y_continuous(limits = c(0, 1200000), breaks = seq(0, 12000000, 200000),label=scales::comma)+
  theme_bw() +
  labs(
  title = "Cell Growth Over Time",
  x = "Day",
  y = "Cells per Cm2",
  fill = "")+
  geom_bar(width=.6, stat = "identity", position = position_dodge(.7))+
  geom_errorbar( aes(x=day, ymin=mean-sd, ymax=mean+sd), width=0.4, colour="black", alpha=0.7, size=.5, position = position_dodge(width = 0.7))+
  facet_wrap(~factor(treatment, levels = treatment_order))+
   geom_point(data = count_data, 
             mapping = aes(x = day, y = cell_cm2), position = position_dodge(width = 0.7))
bar1

#Separated by cell
bar2<-ggplot(counts_sumstats, aes(x=day, y=mean, fill=factor(treatment, level = treatment_order))) + 
  scale_fill_manual(values = setNames(color_map$Color, color_map$Group)) +
  scale_y_continuous(limits = c(0, 1200000), breaks = seq(0, 12000000, 200000),label=scales::comma)+
  theme_bw() +
  labs(
  title = "Cell Growth Over Time",
  x = "Day",
  y = "Cells per Cm2",
  fill = "")+
  geom_bar(width=.6, stat = "identity", position = position_dodge(.7))+
  geom_errorbar( aes(x=day, ymin=mean-sd, ymax=mean+sd), width=0.4, colour="black", alpha=0.7, size=.5, position = position_dodge(width = 0.7))+
  facet_wrap(~cell)+
  geom_point(data = count_data, 
             mapping = aes(x = day, y = cell_cm2), position = position_dodge(width = 0.7))
bar2


```

 
```{r simple line graph}

#sperated by treatment
line1<- ggplot(counts_sumstats, aes(x=day, y=mean, color=cell, group=cell)) + 
  scale_color_manual(values = setNames(color_map$Color, color_map$Group))+
  scale_y_continuous(limits = c(0, 1300000), breaks = seq(0, 12000000, 200000),label=scales::comma)+
  geom_line(size=1)+
  geom_point()+
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.5, size=1,
                 position=position_dodge(0.05))+
  theme_bw() +
  labs(
  title = "Cell Growth Over Time",
  x = "Day",
  y = "Cells per Cm2",
  fill = "" )+
  facet_wrap(~factor(treatment, level = treatment_order), nrow=1)
line1

#sperated by cell
line2<- ggplot(counts_sumstats, aes(x=day, y=mean, color=treatment, group=treatment)) +
  scale_color_manual(values = setNames(color_map$Color, color_map$Group))+
  scale_y_continuous(limits = c(0, 1300000), breaks = seq(0, 12000000, 200000),label=scales::comma)+
  geom_line(size=1)+
  geom_point()+
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.5, size=1,
                 position=position_dodge(0.05))+
  theme_bw() +
  ggtitle("Cell Growth Over Time") +
  xlab("Day") +
  ylab("Cells per Cm2") +
  facet_wrap(~cell)
line2


```


```{r best fit line graph}
#sperated by treatment
linebf1<- ggplot(count_data, aes(x=day, y=cell_cm2, fill=cell, group=cell))+
  scale_fill_manual(values = setNames(color_map$Color, color_map$Group))+
  scale_y_continuous(limits = c(0, 1200000), breaks = seq(0, 12000000, 200000),label=scales::comma)+
  geom_point(shape=21) +
  geom_smooth(method = "loess",
              color="black",
              size=0.5)+
  theme_bw() +
  labs(
  title = "Cell Growth Over Time",
  x = "Day",
  y = "Cells per Cm2",
  fill = "" )+
  facet_wrap(~factor(treatment, levels=c('CTRL','SAG','VIS','RA', 'RA&SAG', 'RA&VIS')),nrow = 1)
linebf1

#sperated by cell
linebf2<- ggplot(count_data, aes(x=day, y=cell_cm2, fill=treatment, group=treatment))+
  scale_fill_manual(values = setNames(color_map$Color, color_map$Group))+
  scale_y_continuous(limits = c(0, 1200000), breaks = seq(0, 12000000, 200000),label=scales::comma)+
  geom_point(shape=21) +
  geom_smooth(method = "loess",
              color="black",
              size=0.5)+
  theme_bw() +
  labs(
  title = "Cell Growth Over Time",
  x = "Day",
  y = "Cells per Cm2",
  fill = "" )+
  facet_wrap(~cell)
linebf2

```

```{r create subfolder for individual treatment graphs}
# Define the path and folder name
folder_path <- "03_plots/cell_growth"

# Create the folder
if (!dir.exists(folder_path)) {
  dir.create(folder_path, recursive = TRUE)
  message("Folder created at: ", folder_path)
} else {
  message("Folder already exists at: ", folder_path)
}
```


```{r save}
#bargraphs
ggsave(bar1, path = "03_plots/cell_growth",
      filename = str_c("bar1-",  format(Sys.Date(), "%Y-%m-%d"), format(Sys.time(), "_%H%M"), ".pdf"),
       height = 3, width = 6, units = "in")

ggsave(bar2, path = "03_plots/cell_growth",
      filename = str_c("bar2-",  format(Sys.Date(), "%Y-%m-%d"), format(Sys.time(), "_%H%M"), ".pdf"),
       height = 3, width = 6, units = "in")

#line graph 
ggsave(line1, path = "03_plots/cell_growth",
      filename = str_c("line1-",  format(Sys.Date(), "%Y-%m-%d"), format(Sys.time(), "_%H%M"), ".pdf"),
       height = 3, width = 6, units = "in")

ggsave(line2, path = "03_plots/cell_growth",
      filename = str_c("line2-",  format(Sys.Date(), "%Y-%m-%d"), format(Sys.time(), "_%H%M"), ".pdf"),
       height = 3, width = 6, units = "in")


#best fit graph 
ggsave(linebf1, path = "03_plots/cell_growth",
      filename = str_c("linebf1-",  format(Sys.Date(), "%Y%m%d"), format(Sys.time(), "-%H%M"), ".pdf"),
       height = 3, width = 6, units = "in")

ggsave(linebf2, path = "03_plots/cell_growth",
      filename = str_c("linebf2-",  format(Sys.Date(), "%Y%m%d"), format(Sys.time(), "-%H%M"), ".pdf"),
       height = 3, width = 6, units = "in")
```



# Step3: Check  Assumptions
```{r variance and normality }

library(car) 

# Data frame to store results
assumption_results <- data.frame(
  Day = character(),
  Treatment1 = character(),
  Treatment2 = character(),
  Cell1 = character(),
  Cell2 = character(),
  Comparison_Group = character(),
  Control_Shapiro_PValue = numeric(),
  Treatment_Shapiro_PValue = numeric(),
  Normality = character(),
  Levene_PValue = numeric(),
  Variance_equal = logical(),
  stringsAsFactors = FALSE
)

# Specify control label and cells for comparison
control_label <- "CTRL"
control_cell <- "S3D21"
treatment_cell <- "S3T21"

# Loop  days to run tests
for (day_val in unique(count_data$day)) {
  
  # Filter  for  current day
  day_data <- count_data %>%
    filter(day == day_val) %>%
    select(cell, treatment, cell_cm2) %>%
    na.omit()
  
  # Loop  cell types to compare each treatment to control
  for (cell_type in unique(day_data$cell)) {
    cell_data <- day_data %>% filter(cell == cell_type)
    
    # Filter control
    control_data <- cell_data %>% filter(treatment == control_label)
    
    # Within each cell type loop through treatments for comparing against control
    for (treatment_val in unique(cell_data$treatment)) {
      if (treatment_val != control_label) {
        
        # Filter for treatment group 
        treatment_data <- cell_data %>% filter(treatment == treatment_val)
        
        # Combine control and treatment data
        comparison_data <- bind_rows(control_data, treatment_data)
        
        # Run tests if at least two groups (control and treatment)
        if (n_distinct(comparison_data$treatment) >= 2) {
          
          # Shapiro-Wilk test for normality within each group, checking for identical values
          control_shapiro <- if (length(unique(control_data$cell_cm2)) > 1) {
            shapiro.test(control_data$cell_cm2)$p.value
          } else {
            NA  # No variability, can't run Shapiro-Wilk
          }
          
          treatment_shapiro <- if (length(unique(treatment_data$cell_cm2)) > 1) {
            shapiro.test(treatment_data$cell_cm2)$p.value
          } else {
            NA
          }
          
          # Determine normality based on available Shapiro results
          normality <- if (!is.na(control_shapiro) && !is.na(treatment_shapiro)) {
            ifelse(control_shapiro > 0.05 & treatment_shapiro > 0.05, "Normal", "Not Normal")
          } else {
            "Indeterminate"
          }
          
          # Levene’s test for homogeneity of variance 
          levene_test <- leveneTest(cell_cm2 ~ treatment, data = comparison_data)
          variance_equal <- levene_test$`Pr(>F)`[1] > 0.05  # TRUE if variances are equal
          
          # Store results for normality and variance tests in a single entry
          assumption_results <- assumption_results %>%
            bind_rows(data.frame(
              Day = day_val,
              Treatment1 = control_label,
              Treatment2 = treatment_val,
              Cell1 = cell_type,
              Cell2 = "NA",
              Comparison_Group = paste(cell_type, "--", treatment_val, "vs", control_label),
              Control_Shapiro_PValue = control_shapiro,
              Treatment_Shapiro_PValue = treatment_shapiro,
              Normality = normality,
              Levene_PValue = levene_test$`Pr(>F)`[1],
              Variance_equal = variance_equal
            ))
        }
      }
    }
  }
  
  # Loop through treatments to compare cell types within that treatment
  for (treatment_val in unique(day_data$treatment)) {
    treatment_data <- day_data %>% filter(treatment == treatment_val)
    
    # Run tests if there are at least two unique cell types
    if (n_distinct(treatment_data$cell) >= 2) {
      
      # Filter data for the control and treatment cell types
      control_data <- treatment_data %>% filter(cell == control_cell)
      treatment_data <- treatment_data %>% filter(cell == treatment_cell)
      
      # Combine control and treatment data
      comparison_data <- bind_rows(control_data, treatment_data)
        
      control_shapiro <- if (length(unique(control_data$cell_cm2)) > 1) {
        shapiro.test(control_data$cell_cm2)$p.value
      } else {
        NA
      }
      
      treatment_shapiro <- if (length(unique(treatment_data$cell_cm2)) > 1) {
        shapiro.test(treatment_data$cell_cm2)$p.value
      } else {
        NA
      }
      
      normality <- if (!is.na(control_shapiro) && !is.na(treatment_shapiro)) {
        ifelse(control_shapiro > 0.05 & treatment_shapiro > 0.05, "Normal", "Not Normal")
      } else {
        "Indeterminate"
      }

      levene_test <- leveneTest(cell_cm2 ~ cell, data = comparison_data)
      variance_equal <- levene_test$`Pr(>F)`[1] > 0.05  # TRUE if variances are equal

      assumption_results <- assumption_results %>%
        bind_rows(data.frame(
          Day = day_val,
          Treatment1 = treatment_val,
          Treatment2 = "NA",
          Cell1 = control_cell,
          Cell2 = treatment_cell,
          Comparison_Group = paste(treatment_val, "--", treatment_cell, "vs", control_cell),
          Control_Shapiro_PValue = control_shapiro,  
          Treatment_Shapiro_PValue = treatment_shapiro,
          Normality = normality,
          Levene_PValue = levene_test$`Pr(>F)`[1],
          Variance_equal = variance_equal
        ))
    }
  }
}

# View the assumption test results
print(assumption_results)

```

```{r save assumptions }
write.csv(assumption_results,"02_tables/cell_growth_assumptions.csv")

```



# STEP3: Run Statistical Analysis
Common tests in one loop
1. Comparison of T21 vs D21 for each treatment for each day
2. Comparision of treatment vs control for each cell type for each day 

```{r comparison }
results_cell <- data.frame(
  Day = character(),
  Treatment = character(),
  Paired_TTest_PValue = numeric(),
  Kruskal_Wallis_PValue = numeric(),
  Welch_TTest_PValue = numeric(),
  Wilcoxon_PValue = numeric(),
  stringsAsFactors = FALSE
)

results_treatment <- data.frame(
  Day = character(),
  Cell = character(),
  Compared_Treatment = character(),  
  Paired_TTest_PValue = numeric(),
  Kruskal_Wallis_PValue = numeric(),
  Welch_TTest_PValue = numeric(),
  Wilcoxon_PValue = numeric(),
  stringsAsFactors = FALSE
)

# Specify control group
control_group <- "CTRL"  
treatment_groups <- unique(count_data$treatment[count_data$treatment != control_group])

# Outer loop: Compare each treatment to the control group
for (treatment_val in treatment_groups) {
  
  # Subset data for the current treatment and control
  subset_data <- count_data %>%
    filter(treatment %in% c(control_group, treatment_val))

  # Inner Loop 1: Compare cell types within a treatment and day for the cell_cm2 measurement
  for (day_val in unique(subset_data$day)) {
    for (current_treatment in unique(subset_data$treatment)) {
      
      # Filter data for current day and treatment
      data_day_treatment <- subset_data %>%
        filter(day == day_val, treatment == current_treatment)
      
      # Initialize p-values as NA
      wilcox_p <- NA
      welch_p <- NA
      paired_p <- NA
      kruskal_p <- NA
      
      # Perform tests if there are at least two cell types
      if (length(unique(data_day_treatment$cell)) > 1) {
        
        # Kruskal-Wallis test
        kruskal_test <- kruskal.test(cell_cm2 ~ cell, data = data_day_treatment)
        kruskal_p <- kruskal_test$p.value

        # Run tests if exactly two cell types are present
        if (length(unique(data_day_treatment$cell)) == 2) {
          
          # Wilcoxon test with exact = FALSE to handle ties
          wilcox_test <- tryCatch({
            wilcox.test(cell_cm2 ~ cell, data = data_day_treatment, exact = FALSE)
          }, error = function(e) NA)
          wilcox_p <- if (!is.na(wilcox_test)) wilcox_test$p.value else NA

          # Welch t-test only if data isn't constant
          if (var(data_day_treatment$cell_cm2) > 0) {
            welch_test <- t.test(cell_cm2 ~ cell, data = data_day_treatment, var.equal = FALSE)
            welch_p <- welch_test$p.value
            
            # Paired t-test if matching subjects are found and data isn’t constant
            if (!any(duplicated(data_day_treatment$subject))) {
              paired_test <- t.test(cell_cm2 ~ cell, data = data_day_treatment, paired = TRUE)
              paired_p <- paired_test$p.value
            }
          }
        }
        
        # Store results for cell type comparisons
        results_cell <- rbind(results_cell, data.frame(
          Day = day_val,
          Treatment = current_treatment,
          Paired_TTest_PValue = paired_p,
          Kruskal_Wallis_PValue = kruskal_p,
          Welch_TTest_PValue = welch_p,
          Wilcoxon_PValue = wilcox_p
        ))
      }
    }
  }
  
  # Inner Loop 2: Compare treatments within each cell type and day for the cell_cm2 measurement
  for (day_val in unique(subset_data$day)) {
    for (cell_val in unique(subset_data$cell)) {
      
      # Filter data for current day and cell type
      data_day_cell <- subset_data %>%
        filter(day == day_val, cell == cell_val)
      
      # Initialize p-values as NA
      wilcox_p <- NA
      welch_p <- NA
      paired_p <- NA
      kruskal_p <- NA
      
      # Perform tests only if there are at least two treatments
      if (length(unique(data_day_cell$treatment)) > 1) {
        
        # Kruskal-Wallis test
        kruskal_test <- kruskal.test(cell_cm2 ~ treatment, data = data_day_cell)
        kruskal_p <- kruskal_test$p.value

        # Run tests if exactly two treatments are present
        if (length(unique(data_day_cell$treatment)) == 2) {
          
          # Wilcoxon test with exact = FALSE to handle ties
          wilcox_test <- tryCatch({
            wilcox.test(cell_cm2 ~ treatment, data = data_day_cell, exact = FALSE)
          }, error = function(e) NA)
          wilcox_p <- if (!is.na(wilcox_test)) wilcox_test$p.value else NA
          
          # Welch t-test only if data isn't constant
          if (var(data_day_cell$cell_cm2) > 0) {
            welch_test <- t.test(cell_cm2 ~ treatment, data = data_day_cell, var.equal = FALSE)
            welch_p <- welch_test$p.value
            
            # Paired t-test if matching subjects are found and data isn’t constant
             if (!any(duplicated(data_day_cell$subject))) {
              paired_test <- t.test(cell_cm2 ~ treatment, data = data_day_cell, paired = TRUE)
              paired_p <- paired_test$p.value
            }
          }
        }
        
        # Store results
        results_treatment <- rbind(results_treatment, data.frame(
          Day = day_val,
          Cell = cell_val,
          Compared_Treatment = paste(treatment_val, "vs", control_group),
          Paired_TTest_PValue = paired_p,
          Kruskal_Wallis_PValue = kruskal_p,
          Welch_TTest_PValue = welch_p,
          Wilcoxon_PValue = wilcox_p
        ))
      }
    }
  }
}

print(results_cell)
print(results_treatment)
 
```


```{r save stats }
write.csv(results_cell,"02_tables/cell_growth_cell_stats.csv")
write.csv(results_treatment,"02_tables/cell_growth_treatment_stats.csv")
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
